Typed Lua: An Optional Type System for Lua

### 2014-03-10 doc/paper/typedlua.tex
correction on projection types; most of the rewrite of section 2.3 done

@@

Typed Lua includes table types to represent Lua's tables.
Although Typed Lua has just one table type constructor, it has three
different syntaxes for defining the type of a table.
The first syntax defines table types for hash tables;
it is written \texttt{\{t1:t2\}},
and maps to the table type $\{t_{1}:t_{2}\;\cup\;\Nil\}$.
This table type represents a hash table that maps values of type
$t_{1}$ to values of type $t_{2}\;\cup\;\Nil$, as Lua returns
$\Nil$ when we use a non-existing key to index a table.
The second syntax defines table types for lists;
it is written \texttt{\{t\}},
and maps to the table type $\{\Number:t\;\cup\;\Nil\}$.
This table type is just a syntactic sugar to the definition of a
hash table that maps values of type $\Number$ to values of type
$t\;\cup\;\Nil$.
The third syntax defines table types for records;
it is written \texttt{\{s1:t1, ..., sn:tn\}},
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$.
This table type represents a record that maps the literals
$s_{i}$\texttt{, ..., }$s_{n}$ to values of types
$t_{i}$\texttt{, ..., }$t_{n}$, where $i$ may vary from one to $n$.

Next, we show one example of table type to type a hash table:
\begin{verbatim}
    local t:{string:number} = { foo = 1 }  -- OK
    local x:number = t.foo                 -- Error
    local y:number? = t.bar                -- OK
\end{verbatim}

The second line of this example gives a type error for the same
reason that the last line of the previous example gives a type error
too, that is, we are trying to assign a value that can be of two
different types to a variable that accepts values of only one type.
Although the field \texttt{bar} does not exist in \texttt{t}, the last
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.

Next, we show one example of table type to type a list:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday",
      "Tuesday", "Wednesday", "Thursday",
      "Friday", "Saturday" }  -- OK
\end{verbatim}

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the list.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\begin{align*}
\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
{5:\String},\;{6:\String},\;{7:\String}\}
\end{align*}
This inferred table type is a subtype of the type
$\{\Number:\String\;\cup\;\Nil\}$, which is the type that we used to
annotate the variable \texttt{days}.

When we use the inferred table type while initializing a table, the
compiler gives an error message when we try to access an index that
is out of bounds.

It also worth mentioning that we can use the style of the inferred
type to define heterogeneous tuples.

Next, we show one example of table type to type a record:
\begin{verbatim}
    local person:{"firstname":string,
                  "lastname":string} =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

Type inference would infer the very same table type that we used to
annotate the variable \texttt{person}.

Typed Lua includes interfaces as syntactic sugar to table types
that specify records.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous example:
\begin{verbatim}
    interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

Now we can use the type \texttt{Person} as an alias to a table type
that certainly has the fields \texttt{firstname} and \texttt{lastname}.
Next, we show an example of code that uses the type \texttt{Person}:
\begin{verbatim}
    local function greeter (person:Person):string
      return "Hello " .. person.firstname ..
             " " .. person.lastname
    end

    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }

    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }

    print(greeter(user1)) -- OK
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- OK
    print(greeter(user4)) -- Error
\end{verbatim}

This example shows that our optional type system is structural rather
than nominal, that is, it checks the structure of types instead of
their names, and it uses subtyping and consistent-subtyping for
checking types.

Using interfaces we can also define recursive types such as linked
lists:

\begin{verbatim}
    interface Element
      info:number
      next:Element?
    end
\end{verbatim}

We need to explicitly annotate \texttt{next} being of type
\texttt{Element} $\cup\;\Nil$ because in our type system $\Nil$
is not the bottom type, that is, $\Nil$ is subtype of itself only.
Typed Lua has this behavior because in Lua $\Nil$ represents
the absence of an useful value.

>>>>

Assigning to a variable with a projection type is forbidden,
unless the union has been discriminated down to a single tuple,
Unrestricted assignment to these variables would be unsound,
as it could break the dependency relation between the types
in each tuple that is part of the union.

@@

Tables are the main mechanism that Lua has to build data
structures. They are associative arrays where any value
(except {\tt nil}) can be a key, but with language support
for efficiently using tables as tuples, arrays (dense or sparse),
records, modules, and objects. In this section,
we show how Typed Lua encodes tables as arrays, records, tuples,
and plain maps in its type system.

Typed Lua uses the same framework to represent the different
uses that a Lua table has: {\em table types}. A table type
$\{ t_{1}:u_{1}, \ldots, t_{n}:u_{n}\}$ represents a map
from values of type $t_i$ to values of type $u_i$.

The concrete syntax of Typed Lua has syntax for
common table types. One syntax defines table types for
maps: it is written \texttt{\{ t: u \}},
and maps to the table type $\{t:u\}$.
This table type represents a map that maps values of type
$t$ to values of type $u \cup \Nil$.
Another syntax defines table types for arrays:
it is written \texttt{\{ t \}}, and maps to the table type
$\{\Number:t\}$. A third syntax sugar defines
table types for records:
it is written \texttt{\{ s1: t1, ..., sn: tn \}}, where
each $s_i$ is a literal number, string, or boolean, 
and maps to the table type $\{s_{1}:t_{1}, ..., s_{n}:t_{n}\}$,
where each $s_i$ is the corresponding literal type.

The example below shows how we can define a map from
strings to numbers; the dot syntax for field access
in Lua is actually syntactic sugar for indexing a table
with a string literal:

\begin{verbatim}
    local t: { string: number } = { foo = 1 }
    local x: number = t.foo         -- x gets 1
    local y: number = t["bar"]      -- runtime error
\end{verbatim}

When accessing a map, there is always the possibility that
the key is not there. In Lua, accessing a non-existing key
returns {\tt nil}. Typed Lua is stricter, and raises a runtime
error in this case. To get a map with the behavior of standard
Lua tables, the programmer can use an union:

\begin{verbatim}
    local t: { string: number? } = { foo = 1 }
    local x: number = t.foo         -- error
    local y: number = t.bar or 0    -- y gets 0
    local z: number? = t["bar"]     -- z gets nil
\end{verbatim}

Now the Typed Lua compiler will complain about the assignment
on line two. The following example shows how we can declare
an array:
\begin{verbatim}
    local days: { string } = { "Sunday", "Monday",
      "Tuesday", "Wednesday", "Thursday",
      "Friday", "Saturday" }
    local x = days[1]          -- x gets "Sunday"
    local y = days[8]          -- runtime error
\end{verbatim}

Notice that we have the same strictness with missing
elements, unless the type of the elements has {\tt nil}
as a possible value.

If we want to declare a tuple, we can leave the
variable declaration unannotated and let Typed Lua assign
a more specific table type to the variable.
If we remove the annotation in the previous example, 
the compiler assigns the following table type to \texttt{days}:
\begin{align*}
\{{1:\String},\;{2:\String},\;{3:\String},\;{4:\String},\;\\
{5:\String},\;{6:\String},\;{7:\String}\}
\end{align*}

This type is not a subtype of $\{\Number:\String\}$, nor
is $\{\Number:\String\}$ a subtype of $\{\Number:
\String\cup\Nil\}$, because in both cases the subtype
relationship would be unsound. In the first case,
a table type with the same fields as the type above,
plus $8: \Number$, is a subtype of the table type
above, so would also be a subtype of $\{\Number:\String\}$,
which is clearly unsound. In the second case, covariance
in the type of mutable fields is also unsound, for the
same reason as the unsoundness of array covariance.

While the record type above, $\{\Number:\String\}$, and
$\{\Number: \String\cup\Nil\}$ are disjoint, all three
types are valid for the table constructed in the example.
Typed Lua actually assigns different types to a table constructor
expression depending on the context where it is used.

Finally, the next example shows how we can declare a record:

\begin{verbatim}
    local person: { "firstname": string,
                    "lastname": string } =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

We could leave the type annotation out, and Typed Lua would
assign the same type to {\tt person}.

As records get bigger, and types of record fields get more
complicated, writing table types can be unwieldy, so Typed
Lua has {\em interfaces} as syntactic sugar for record types:

\begin{verbatim}
    local interface Person
      firstname: string
      lastname: string
    end
\end{verbatim}

The declaration above declares {\tt Person} as an alias to
the record type $\{$``firstname": $\String$,
 ``lastname": $\String\}$ in
the remainder of the current scope. We can now use {\tt Person} in
type declarations:

\begin{verbatim}
    local function greet(person: Person)
      return "Hello, " .. person.firstname ..
             " " .. person.lastname
    end

    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }
    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }

    print(greeter(user1)) -- Hello, Lewis Reed
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- Hello, Lou Reed
    print(greeter(user4)) -- Error
\end{verbatim}

If our record type has fields that can be {\tt nil}, we need
to use an explicit type declaration when declaring a
variable of this record type, as the following example shows:

\begin{verbatim}
    local interface Person
      firstname: string
      middlename: string?
      lastname: string
    end

    local user1: Person = { firstname = "Lewis",
                            middlename = "Allan",
                            lastname = "Reed" }
    local user2: Person = { lastname = "Reed",
                            firstname = "Lou" }
\end{verbatim}

We need an explicit type declaration because neither of
the types that the Typed Lua compiler assigns to the
table constructors above is a subtype of $\{$``firstname":
$\String$, ``middlename": $\String \cup \Nil$, ``lastname":
$\String\}$, the type that {\tt Person} describes.

We can also use interfaces to define recursive types:

\begin{verbatim}
    local interface Element
      info: number
      next: Element?
    end
\end{verbatim}

It is common in Lua programs to build a record incrementally,
starting with an empty table, as in the following example:

\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

Ideally, we want the type of {\tt person} to change as the
table gets built, from $\{\}$ to $\{$``firstname": $\String\}$
and finally to $\{$``firstname": $\String$, ``lastname":
 $\String\}$. This is tricker than the type change introduced
by assignment that we saw in Section 2.2, as what is changing
is not just the type of the variable {\tt person} but the
type of the value that {\tt person} points to.
